% =============================================================================
% Abstract — conservative, evidence-aligned framing
% Drop-in include: \input{sections/abstract}
% Target: <= 250 words
% =============================================================================
\begin{abstract}
RL post-training of language models is now shaped by a small set of
algorithms (PPO, GRPO, DPO) and a growing set of frameworks (TRL,
\tinker{}, OpenRLHF, veRL), yet it remains unclear which empirical conclusions
transfer across implementations, model families, and scales. We present
\ours{}, a benchmark spanning 79 runs across 7 RL libraries, 5 model
families, and 0.6B--$\sim$1T parameters on GSM8K, MATH-500, HumanEval, and
synthetic tool-use tasks. The strongest evidence in the current release comes
from short-horizon GSM8K training dynamics; held-out evaluation and non-math
coverage remain narrower.

Three conclusions are supported conservatively. First, Zero-Variance Fraction
(ZVF)---the share of GRPO groups with identical rewards---tracks when binary
outcome rewards stop producing within-group learning signal. Because ZVF is
mechanically coupled to reward sparsity, group size, and baseline accuracy, we
use it as a descriptive diagnostic rather than a standalone causal predictor.
Second, trainability in our current sweeps varies substantially with
initialization and rollout regime: instruction-tuned checkpoints were
generally easier to optimize than comparable base models, and intermediate
group sizes often behaved better than the smallest or largest settings we
tested, but we do not claim a universal optimal group size. Third, PPO/GRPO
rankings and frontier-model stability are heterogeneous across model families
and end-to-end stacks; our single-seed API runs should be read as case studies,
not universal laws. A separate held-out GSM8K control shows that the mean
Qwen3-8B GRPO delta over the base model is small and not statistically
significant (83.3\% vs.\ 82.0\%, $p{=}0.26$), while tool-use and code results
remain limited by sparse rewards and custom evaluation.

We release code, logs, and checkpoints at
\url{https://anonymous.4open.science/r/tinker-rl-lab}.
\end{abstract}
