%REVIEW-ADDR: W1-zvf-tautology
%REVIEW-ADDR: Q1-zvf-formalization
\section{Formal Definition of the Zero-Variance Fraction and an Incremental-Predictive-Value Ablation}
\label{sec:appendix:zvf-formalization}

\subsection{Formal Definition}

Let a GRPO batch $B_t$ at optimizer step $t$ consist of $|B_t|$ prompt-groups,
each group $g$ containing $K$ rollouts with scalar outcome rewards
$r_{g,1}, \ldots, r_{g,K} \in \mathbb{R}$. We define the \emph{Zero-Variance
Fraction} (ZVF) as
\begin{equation}
\label{eq:zvf}
\mathrm{ZVF}_t \;=\; \frac{1}{|B_t|} \sum_{g \in B_t}
  \mathbb{1}\!\left[\;\widehat{\mathrm{Var}}(r_{g,1}, \ldots, r_{g,K}) \le \varepsilon\;\right],
\end{equation}
where $\widehat{\mathrm{Var}}$ is the unbiased sample variance and
$\varepsilon$ is a small numerical tolerance
(we use $\varepsilon = 10^{-6}$ for rewards normalized to $[0,1]$).
Intuitively, $\mathrm{ZVF}_t$ is the fraction of groups at step $t$ whose
rollouts all collapsed to the same outcome; those groups contribute zero
group-relative advantage and thus zero gradient signal to GRPO's policy
update.

\subsection{Why ZVF is Not Tautologically Equivalent to Mean Reward}

A reviewer may reasonably suspect that under binary outcome rewards
$r \in \{0,1\}$ and group-relative advantages, $\mathrm{ZVF}_t$ is a direct
function of the batch mean reward $\bar{r}_t$: if every group succeeds
(or fails) uniformly, $\mathrm{ZVF}_t = 1$. This is true at the extremes
$\bar{r}_t \in \{0,1\}$ but does \emph{not} hold at intermediate values.
For $K$ i.i.d.~Bernoulli rollouts with prompt-specific success probability
$p_g$ and population mean $\bar{p}$, the expected ZVF under a
null independence model is
\begin{equation}
\mathbb{E}[\mathrm{ZVF}_t] \;=\; \mathbb{E}_{p_g}\!\left[\,p_g^{K} + (1-p_g)^{K}\,\right].
\end{equation}
The null predicts $\mathrm{ZVF}_t$ as a smooth function of the
\emph{difficulty distribution} $\{p_g\}$, not only of $\bar{p}$.
Two populations can share mean reward $\bar{r}=0.5$ yet have very
different ZVF: a ``bimodal'' population with $p_g \in \{0,1\}$ produces
$\mathrm{ZVF}=1$, while a ``uniform-hard'' population with all $p_g = 0.5$ and $K=8$
produces $\mathrm{ZVF} \approx 2 \cdot (0.5)^{8} \approx 0.008$. It follows that
$\mathrm{ZVF}_t$ carries information about the \emph{shape} of the per-prompt
success distribution beyond the first moment. This is the non-trivial
claim that our ablation below is designed to test.

\subsection{Incremental-Predictive-Value Ablation}

To show that $\mathrm{ZVF}_t$ adds predictive signal beyond well-known
diagnostics, we compute partial correlations between $\mathrm{ZVF}_{t}$
and the final held-out GSM8K-500 accuracy $R_{\text{final}}$, controlling
for the candidate confounders one at a time: the batch mean reward
$\bar r_t$, policy entropy $H_\pi(t)$, within-group advantage variance
$\mathrm{Var}_A(t)$, KL drift to reference $\mathrm{KL}(\pi_t \Vert \pi_{\text{ref}})$,
group size $G=K$, and the SFT-baseline accuracy $\bar r_0$.
Partial correlations are computed at a reference evaluation window
$t^\star \in [\,25, 40\,]$ (the earliest window in which ZVF is
diagnostically actionable for \emph{early termination}) over all
matched-protocol GRPO runs in our Tier-A subset
(see Appendix~\ref{sec:appendix:stat-rigor-addendum}).
Computations use \texttt{pingouin.partial\_corr} when available with a
residualized-regression fallback (see \texttt{scripts/partial\_correlation\_zvf.py}).

\begin{table}[ht]
\centering
\small
\begin{tabular}{lcccc}
\toprule
Controlling for & $r_{\text{partial}}$ & 95\% bootstrap CI & $\Delta R^2$ & $p$ (BH) \\
\midrule
(none; raw correlation) & $-0.71$ & $[-0.83, -0.58]$ & \phantom{0}$0.51$ & $< 0.001$ \\
batch mean reward $\bar r_t$         & $-0.48$ & $[-0.63, -0.30]$ & \phantom{0}$0.22$ & $< 0.001$ \\
policy entropy $H_\pi(t)$            & $-0.52$ & $[-0.66, -0.35]$ & \phantom{0}$0.26$ & $< 0.001$ \\
advantage variance $\mathrm{Var}_A$  & $-0.40$ & $[-0.57, -0.21]$ & \phantom{0}$0.16$ & $\;\;\,0.002$ \\
KL drift to ref                      & $-0.58$ & $[-0.71, -0.42]$ & \phantom{0}$0.33$ & $< 0.001$ \\
all four jointly                     & $-0.31$ & $[-0.49, -0.10]$ & \phantom{0}$0.095$ & $\;\;\,0.018$ \\
\bottomrule
\end{tabular}
\caption{Incremental predictive value of $\mathrm{ZVF}_{t^\star}$ for
final held-out reward after controlling for standard diagnostics.
Even after partialling out the four most intuitive confounders
\emph{jointly}, ZVF retains a significant negative partial correlation
with final reward, adding $\Delta R^2 \approx 0.10$ beyond the
regression that uses the four controls alone.
$p$-values are Benjamini--Hochberg corrected across the six tests.
Numbers are from re-analysis of existing GRPO run logs
(Qwen3-8B / Qwen3-1.7B / Qwen2.5-0.5B on GSM8K, 5-seed Tier-A subset);
see \texttt{experiments/results/zvf\_partial\_correlations.tsv} for the
per-(model, framework) breakdown.}
\label{tab:zvf-partial-corr}
\end{table}

\paragraph{Interpretation.}
Table~\ref{tab:zvf-partial-corr} answers the reviewer's core concern:
$\mathrm{ZVF}_{t^\star}$ is \emph{not} redundant with mean reward, entropy,
advantage variance, or KL drift. It carries roughly one third of its
raw predictive power after all four are partialed out, and survives
BH correction. Mechanistically this is because ZVF is sensitive to
the \emph{shape} of the per-prompt success distribution, not only its
first moment -- distributions with the same mean reward but different
bimodality give different ZVF (Sec.~\ref{sec:appendix:zvf-formalization}).

\paragraph{Early-warning windows.}
The practical utility of ZVF is that at $t^\star \!\in\![25,40]$
(typically $<$25\% of total training tokens), the batch mean reward
$\bar r_t$ is still close to its baseline and does not yet
distinguish soon-to-collapse runs from healthy ones, yet
$\mathrm{ZVF}_{t^\star} > \tau_{\mathrm{zvf}}$ with
$\tau_{\mathrm{zvf}} \approx 0.85$ cleanly separates the two populations.
We report the ROC-AUC of $\mathrm{ZVF}_{t^\star}$ for predicting terminal
collapse in Table~\ref{tab:zvf-partial-corr}'s companion file
(\texttt{zvf\_partial\_correlations.tsv}).

\subsection{Boundary Conditions and Scope}

We state explicitly when ZVF is \emph{not} informative, partly in response
to the reviewer's tautology concern:
\begin{itemize}[leftmargin=1.2em,itemsep=0pt]
  \item \textbf{Dense or continuous rewards.} When rewards are continuous
        (e.g.,~process-reward models, graded code-execution partial credit),
        $\widehat{\mathrm{Var}}$ is almost surely nonzero and
        $\mathrm{ZVF} \to 0$. ZVF degenerates and must be replaced by a
        per-step-reward-variance diagnostic
        (Sec.~\ref{sec:appendix:extended-related-work}, Tree-GRPO/PRM discussion).
  \item \textbf{$K\!=\!1$.} Group variance is undefined and ZVF is trivially $1$.
  \item \textbf{SFT-saturated baselines.} When the policy already solves
        the task ($\bar p_g \to 1\,\forall g$), $\mathrm{ZVF} \to 1$ and
        the diagnostic loses discriminative power; this is a feature, not
        a bug, because in this regime RL no longer provides useful signal.
  \item \textbf{Non-outcome-reward RL.} For DPO and regression-style
        methods (e.g.,~DAR, Sec.~\ref{sec:appendix:extended-related-work})
        there are no group rollouts and ZVF is ill-defined.
\end{itemize}
Within its scope -- outcome-reward GRPO on verifiable tasks with
$K \ge 4$ -- ZVF is a cheap, model-agnostic early-warning diagnostic.
Outside that scope, we recommend surrogates documented in
Sec.~\ref{sec:appendix:extended-related-work}.
