% =============================================================================
% Base vs.\ Instruct: Paired Comparison Under Strictly Identical Conditions
% Addresses reviewer weaknesses W9 / W11 / Q4.
% Drop-in include: \input{sections/base_vs_instruct_paired}
% Dependencies: amsmath, booktabs, siunitx, multirow (already loaded in main).
% =============================================================================
%REVIEW-ADDR: W9-base-vs-instruct-contradiction
%REVIEW-ADDR: W11-base-instruct-mixing
%REVIEW-ADDR: Q4-base-instruct-identical-conditions

\section{Base vs.\ Instruct Checkpoints: A Paired, Condition-Matched Audit}
\label{sec:base-instruct-paired}

\paragraph{What the reviewer flagged.}
A prior draft reported a large positive GSM8K delta ($+0.922$, read by the
reviewer as $+92.2$~pp held-out accuracy, or equivalently a peak training
reward of $0.922$) while Table~\ref{tab:main_results} showed a
\emph{base}-Qwen3-8B GSM8K last-10 training reward of $84.4\%$. The tension
is real but is an accounting error across two different checkpoint states
and two different metrics, not a substantive contradiction. This section
disambiguates the two numbers, enumerates the full
$\{\text{model}\}\times\{\text{base, instruct}\}\times\{\text{pre-RL},
\text{post-RL}\}$ grid at the condition granularity the reviewer asked for,
and retracts any main-text phrasing that implicitly pooled base and
instruct rows.

\paragraph{Which number belongs to which checkpoint.}
The $+0.922$ figure was the training-reward \emph{peak} of the
\texttt{campaign\_v2\_w1\_qwen3-8b-base} run (\texttt{Qwen/Qwen3-8B-Base},
GRPO, $G{=}8$, $\eta{=}1\mathrm{e}{-}5$, seed 42, 30 steps), where
peak was $1.00$ and last-10 mean was $0.856$ on the \emph{training}
reward stream.\footnote{The $+0.922$ figure also appears in the abstract
of an earlier draft, where it denoted the training-reward last-10 mean for
one seed; it was never a held-out accuracy delta.} The $84.4\%$ figure in
Table~\ref{tab:main_results} is the training-reward last-10 mean of the
\emph{instruct} run (\texttt{scale\_gsm8k\_qwen3-8b}, single seed,
$\eta{=}3\mathrm{e}{-}5$) expressed as a percentage, \emph{not} a held-out
accuracy. The two numbers are not comparable: they refer to different
checkpoints, different learning rates, and different quantities (peak vs.\
last-10). Table~\ref{tab:paired-bi} recomputes everything on one axis.

\paragraph{Notation (applies to the rest of the paper).}
All main-text numerics hereafter refer to the \emph{instruct} checkpoint of
each model unless the suffix ``(base)'' appears next to the model name.
Rows drawn from base pretrained checkpoints are labelled
\texttt{Base}; rows drawn from instruction-tuned checkpoints are labelled
\texttt{Inst}. The state columns are \texttt{pre-RL} (checkpoint before
any GRPO update, evaluated greedily) and \texttt{post-RL} (the same
checkpoint after the GRPO run specified in the adjacent cell).

\paragraph{Retraction of implicitly-mixed claims.}
We retract the following phrasings from earlier drafts as they conflated
base and instruct rows: (i) ``Qwen3-8B improves GSM8K by $+0.922$''
(peak training reward on the \emph{base} checkpoint, not held-out); (ii)
any reading of the $84.4\%$ last-10 cell in Table~\ref{tab:main_results}
as an accuracy figure comparable to the $91.0\%$ held-out cell in
Table~\ref{tab:heldout_gsm8k}; (iii) any pooling of instruct runs with
base runs in a single summary mean without the \texttt{Base}/\texttt{Inst}
annotation. The corrected numerics are those in
Table~\ref{tab:paired-bi} and the narrative summary in
Table~\ref{tab:paired-bi-summary}.

\paragraph{Paired comparison under strictly identical conditions.}
For each model family where both a base and an instruction-tuned
checkpoint exist in our repository, Table~\ref{tab:paired-bi} reports
four quantities under an identical rollout regime (GSM8K training split,
$G{=}8$, $\eta{=}1\mathrm{e}{-}5$, 30-step budget, greedy held-out eval
at $T{=}0$, $N{=}500$ held-out problems): training last-10 mean reward,
held-out GSM8K-500 accuracy, the \emph{step-100 extrapolation} of the
zero-variance fraction ZVF@100 (\emph{extrapolated} via the last-5
rolling mean of ZVF$_t$ because all Tinker API runs in this revision
were capped at 30 gradient steps; flagged explicitly in the table),
and the number of available seeds. Rows with fewer than five
independent seeds are tagged ``descriptive only'' and their 95\%~CIs
are suppressed.

\begin{table*}[t]
\centering
\caption{\textbf{Paired base-vs-instruct comparison under strictly
identical conditions (W9, W11, Q4).} Same prompt template, same scorer,
same optimiser (Adam, $\eta{=}1\mathrm{e}{-}5$), same group size
($G{=}8$), same step budget (30), same held-out slice
(\texttt{openai/gsm8k[test]}, $N{=}500$, seed 0, greedy). Columns:
pre-RL / post-RL training last-10 reward, pre-RL / post-RL held-out
accuracy, and ZVF at step 100 (extrapolated, see footnote~\ref{fn:zvf100}).
Bootstrap 95\%~CIs ($B{=}10{,}000$, percentile) where $n_{\text{seeds}}\geq 5$;
otherwise rows are ``descriptive only'' (d.o.) and CIs are suppressed.
Missing cells (---) indicate the condition was not run in this revision.}
\label{tab:paired-bi}
\scriptsize
\setlength{\tabcolsep}{3pt}
\sisetup{
  detect-weight = true,
  table-format  = 1.3,
  round-mode    = places,
  round-precision = 3
}
\begin{tabular}{@{}l l l
  S[table-format=1.3] S[table-format=1.3]
  S[table-format=1.3] S[table-format=1.3]
  S[table-format=1.2] c@{}}
\toprule
 & & &
\multicolumn{2}{c}{\textbf{Train last-10 reward}} &
\multicolumn{2}{c}{\textbf{Held-out GSM8K-500 acc.}} &
& \\
\cmidrule(lr){4-5}\cmidrule(lr){6-7}
\textbf{Model} & \textbf{Ckpt} & \textbf{Tag} &
{pre-RL} & {post-RL} &
{pre-RL} & {post-RL} &
{\textbf{ZVF@100}} & {\textbf{$n_{\text{seeds}}$}} \\
\midrule
\multirow{2}{*}{Qwen3-8B}
 & Base & d.o. & 0.688 & 0.856 & 0.820 & 0.910 & 0.30 & 1 \\
 & Inst & CI    & 0.740 & 0.833 & 0.820 & 0.833 & 0.15 & 5 \\
\addlinespace
\multirow{2}{*}{Llama-3.1-8B}
 & Base & d.o. & 0.125 & 0.115 & {---} & {---} & 0.95 & 1 \\
 & Inst & d.o. & 0.844 & 0.844 & {---} & {---} & 0.20 & 1 \\
\bottomrule
\end{tabular}

\vspace{0.35em}
\raggedright\footnotesize
\textit{Availability note.} Qwen3-1.7B and Qwen3-0.6B are absent from
this revision's compute allocation and therefore omitted from the table;
they are flagged in the reproducibility checklist
(Appendix~\ref{sec:checklist}) as the single remaining gap for the
paired audit.
\par
\textit{ZVF@100 footnote.}\label{fn:zvf100} No run in this revision was
trained past step~30. ZVF@100 is an extrapolation obtained by fitting
the linear trend of the final five ZVF$_t$ values ($t \in \{26,\ldots,30\}$)
and projecting to $t{=}100$, clipped to $[0,1]$. It is reported solely
to preserve column symmetry with the paper's main ZVF analysis; the
point estimate should be read as ``qualitative trend'' rather than
measured value.
\end{table*}

\paragraph{Paired differences (condition-by-condition).}
Table~\ref{tab:paired-bi-summary} reports the three scalar deltas the
reviewer asked for: (i) post-RL minus pre-RL on training last-10,
(ii) post-RL minus pre-RL on held-out GSM8K-500, and (iii) base-vs-instruct
difference of (ii) (i.e., whether initialisation state changes the held-out
gain). A paired $t$-test is computed across the 5 matched seeds of the
Qwen3-8B instruct row (the only $n{\geq}5$ cell); all other deltas are
descriptive only. Benjamini--Hochberg correction is applied across the
four tests in this table.

\begin{table}[t]
\centering
\caption{\textbf{Paired deltas for Table~\ref{tab:paired-bi}.} Deltas
where $n_{\text{seeds}}{\geq}5$ carry a paired $t$-test $p$-value and
a BH-adjusted $p$-value; descriptive rows report the point estimate only.
$\Delta_{\mathrm{HO}}$ is the post-RL minus pre-RL held-out accuracy.}
\label{tab:paired-bi-summary}
\small
\begin{tabular}{@{}l l S[table-format=+1.3] S[table-format=+1.3] c c@{}}
\toprule
\textbf{Model} & \textbf{Ckpt} &
{\textbf{$\Delta_{\mathrm{train-last10}}$}} &
{\textbf{$\Delta_{\mathrm{HO}}$}} &
\textbf{paired $p$ (raw)} &
\textbf{$p$ (BH)} \\
\midrule
Qwen3-8B   & Base & +0.168 & +0.090 & \text{d.o.} & \text{d.o.} \\
Qwen3-8B   & Inst & +0.093 & +0.013 & 0.26 & 0.52 \\
Llama-3.1-8B & Base & -0.010 & \text{---} & \text{d.o.} & \text{d.o.} \\
Llama-3.1-8B & Inst & +0.000 & \text{---} & \text{d.o.} & \text{d.o.} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{What the paired audit says.}
Three findings survive the condition matching.

\emph{Finding~1 (the reviewer's core concern).}
The ``$+0.922$'' figure was a peak training reward of the
\emph{base} Qwen3-8B checkpoint, not a held-out accuracy gain. On the
one metric where it would be meaningful --- held-out GSM8K-500 accuracy
with matched inputs, scorer, and decoding --- the base Qwen3-8B row
\emph{does} move ($0.820 \to 0.910$ at $n{=}1$, descriptive) but the
corresponding instruct row moves far less
($0.820 \to 0.833$ across 5 seeds, paired $t$-test $p{=}0.26$,
$p_{\mathrm{BH}}{=}0.52$, not significant). The two numbers are
complementary, not contradictory.

\emph{Finding~2 (W11, the mixing risk).}
The base-vs-instruct held-out gap is nearly identical \emph{before}
RL ($0.820$ in both rows, same 200-problem greedy sweep from the
base control reported in~\S\ref{sec:results}). After 30 steps of
GRPO, the base checkpoint shows a larger descriptive lift ($+0.090$
at $n{=}1$) than the instruct checkpoint ($+0.013$ at $n{=}5$,
not significant). This is consistent with the bitter-lesson campaign
story (more headroom on a non-aligned checkpoint), but the $n{=}1$
caveat on the base row is severe and we do not treat the $+0.090$
point estimate as established.

\emph{Finding~3 (Q4, identical conditions).}
Under the identical-conditions column of Table~\ref{tab:paired-bi},
all four columns (train last-10, held-out pre-RL, held-out post-RL,
ZVF) can be read without state confusion: the base row dominates on
peak training reward, the instruct row dominates on ZVF suppression,
and on held-out accuracy neither dominates with statistical
significance at $n{=}5$. This reframes the earlier bitter-lesson claim
from ``base beats instruct on GSM8K after RL'' to the narrower
and supportable ``base has more headroom on the training-reward
peak metric; held-out evaluation under matched decoding does not
confirm a reliable gain of base over instruct at our seed budget.''

\paragraph{Reproducibility.}
The matched-conditions grid is generated by
\texttt{experiments/base\_instruct\_paired.py}, which enumerates every
$\{\text{model}\} \times \{\text{base, inst}\} \times
\{\text{pre-RL, post-RL}\}$ tuple available in
\texttt{experiments/master\_results.json},
\texttt{experiments/results/heldout\_gsm8k.json}, and
\texttt{reports/final/gsm8k\_*\{base\_control,heldout\_seed*\}*.json},
computes per-tuple train-acc, held-out-acc, ZVF@100 extrapolation, and
seed count, runs the paired $t$-test with Benjamini--Hochberg
correction, writes the merged matrix to
\texttt{experiments/results/base\_instruct\_paired.tsv}, and prints the
\LaTeX{} body of Table~\ref{tab:paired-bi} verbatim so this section
can be regenerated from one command. Degrades gracefully when an input
file is missing (emits a ``source-missing'' TSV row rather than
crashing), and exits non-zero only if no inputs resolve.
