%REVIEW-ADDR: W6-framework-gap-contradiction
%REVIEW-ADDR: Q3-framework-exact-configs
\section{Per-Framework Configuration Dumps and Scope of the ``Matched-Config'' Comparison}
\label{sec:appendix:framework-configs}

\subsection{Retraction of the ``Byte-Identical'' Framing}

The main text used the phrase ``byte-identical training configs'' to
describe the framework comparison. A reviewer correctly identified
that this is in tension with our own acknowledgement that the
Tinker-managed runtime applies ``managed reference-model offload and
tuned rollout defaults''. We retract the ``byte-identical'' phrasing
and replace it with the more precise \emph{matched-configuration}
protocol described below.

\subsection{What Was Held Constant Across Frameworks}

\begin{itemize}[leftmargin=1.2em,itemsep=1pt]
  \item Base model weights and tokenizer (identical SHA-256 checksum of
        HuggingFace snapshot)
  \item LoRA rank $r$, $\alpha$, and target-module set
  \item Surfaced optimizer family (AdamW) and requested learning rate
  \item Group size $G$ and rollouts per group $K$
  \item Total tokens seen by the optimizer (not total steps)
  \item Seed
  \item Prompt template and response-mask policy (response-only loss)
\end{itemize}

\subsection{What Was \emph{Not} Identical (Framework-Managed)}

\begin{itemize}[leftmargin=1.2em,itemsep=1pt]
  \item Reference-model placement: TRL keeps $\pi_\text{ref}$ on-device;
        Tinker transparently offloads/recomputes; OpenRLHF uses a
        dedicated reference-serving process; veRL uses Ray-managed
        sharded placement.
  \item Rollout batch micro-partitioning and KV-cache reuse across
        rollouts of the same prompt.
  \item Importance-sampling granularity (token-level in TRL/veRL,
        sequence-level in OpenRLHF at the time of writing).
  \item KL-coefficient default: TRL and veRL use $\beta=0.04$,
        OpenRLHF $\beta=0.02$, Tinker $\beta$ is managed (we set the
        surfaced value to $0.04$ but the internal schedule is
        proprietary).
  \item Numerical precision of the reward-broadcast step (fp32 vs
        bf16 in some paths).
\end{itemize}

\subsection{Complete Configuration Dumps}

Full YAML dumps of every hyperparameter the framework exposes are
released at
\texttt{experiments/framework\_config\_dumps/\{trl,tinker,openrlhf,verl\}\_qwen3\_8b\_gsm8k.yaml}
(see also the accompanying
\texttt{experiments/framework\_config\_dumps/README.md} for per-field
commentary and reproduction instructions).
Fields the Tinker API does not expose are marked
\texttt{null\ \#\ managed\_by\_tinker}; readers can verify that, of
the 47 hyperparameter fields we track, 31 are identical across the
four frameworks, 11 are framework-specific but documented, and 5 are
Tinker-managed and therefore not controlled.

\subsection{Measured vs.\ documented rows in the current artifact}

The configuration dumps are released as \emph{surfaced-config artifacts}.
In the current anonymous bundle, the Tinker and TRL framework-gap rows are
backed by measured runs, whereas the OpenRLHF and veRL rows in
\texttt{experiments/results/framework\_comparison.json} are deterministic
dry-run placeholders emitted by the aggregation script when those
frameworks were unavailable in the sandbox. We therefore use the four YAMLs
to document the intended matched-configuration scaffold across all four
frameworks, but we treat only the Tinker-vs.-TRL comparison as empirical
evidence in this artifact.

\begin{table}[ht]
\centering
\small
\begin{tabular}{lcccc}
\toprule
Hyperparameter & TRL & TINKER & OpenRLHF & veRL \\
\midrule
LoRA rank $r$                  & 16   & 16   & 16   & 16   \\
LoRA $\alpha$                  & 32   & 32   & 32   & 32   \\
group size $G$                 & 8    & 8    & 8    & 8    \\
rollouts per group $K$         & 1    & 1    & 1    & 1    \\
max new tokens                 & 384  & 384  & 384  & 384  \\
sampling temperature           & 0.7  & 0.7  & 0.7  & 0.7  \\
KL $\beta$ (surfaced)          & 0.04 & \emph{managed} & 0.02 & 0.04 \\
IS granularity                 & token & token & seq  & token \\
$\pi_\text{ref}$ placement      & on-dev & \emph{managed} & separate & sharded \\
AdamW lr                       & $1\!\times\!10^{-5}$ & $1\!\times\!10^{-5}$ & $1\!\times\!10^{-5}$ & $1\!\times\!10^{-5}$ \\
tokens/optimizer step          & 3072 & \emph{managed} & 3072 & 3072 \\
\bottomrule
\end{tabular}
\caption{Representative subset of the per-framework hyperparameter table
(full version in the YAML dumps). Fields in italics are Tinker-managed
and unavailable to the user. OpenRLHF and veRL entries document the
intended surfaced configuration of the comparison scaffold; in the current
artifact their performance rows are dry-run placeholders rather than new
measured baselines. This table directly addresses the reviewer request for
``exact config dumps so readers can assess fairness''.}
\label{tab:framework-configs}
\end{table}

\subsection{Corrected Scope of the Framework Comparison}

We revise the interpretation of framework-gap results accordingly:
reported differences should be read as ``behavioural differences under
the matched-configuration protocol defined above'', \emph{not} as
``differences due to algorithmic variants alone''. In particular, any
advantage attributed to Tinker includes the benefit of its managed
reference-model offload and rollout-micro-partitioning, which are
genuine engineering contributions rather than algorithmic ones. In the
current artifact, the only measured gap is Tinker versus TRL; OpenRLHF and
veRL remain documented comparison targets whose surfaced configurations are
released, but whose runtime rows are not treated as fresh empirical evidence.
