% =====================================================================
% paper/sections/statistical_rigor_addendum.tex
%
% Statistical Rigor Addendum: evidence-tier partition and survival
% analysis of F1-F5 after excluding single-seed Tinker/API runs from
% inferential tests. Companion artifact: experiments/survival_analysis.py.
%
% Reviewer addressing:
%REVIEW-ADDR: W4-single-seed-tinker
%REVIEW-ADDR: W5-bh-aggregates-singleseed
%REVIEW-ADDR: Q5-f1-f5-survival-5seed
%
% Compiles under paper/main.tex; uses only amsmath / booktabs / siunitx
% (all already loaded) and introduces no new packages.
% =====================================================================

\section{Statistical Rigor Addendum: Evidence Tiers and Survival}
\label{app:stat-rigor-addendum}

This appendix addresses three interrelated reviewer concerns. \emph{(W4)}
A subset of our headline numbers -- in particular the Tinker API
GRPO runs on frontier models -- come from a single seed at $20$--$30$
gradient steps, which provides essentially zero statistical power.
\emph{(W5)} The Benjamini--Hochberg (BH) corrections reported in
Tables~\ref{tab:main_results_stats}--\ref{tab:ppo_grpo_stats} treat
single-seed and multi-seed rows as exchangeable members of a single
family, which inflates the apparent significance of multi-seed
findings by riding on noisier single-seed entries.
\emph{(Q5)} A natural reviewer question is therefore: which of the
headline findings $F_1$--$F_5$ actually survive when we restrict the
analysis to TRL as the single open framework, at $\ge 5$ seeds and
$\ge 100$ steps?

We address (W4), (W5) and (Q5) by (i) partitioning every run in the
release into three evidence tiers, (ii) excluding Tier-C
(single-seed/short-horizon) runs from any inferential test, and
(iii) re-running BH with the restricted $p$-value family. All numbers
in this appendix are produced by
\texttt{experiments/survival\_analysis.py} from the raw run records in
\texttt{experiments/results/*.jsonl},
\texttt{experiments/master\_results.json} and
\texttt{experiments/master\_results.csv}. The script also writes a
machine-readable \texttt{experiments/results/survival\_analysis.tsv}.

\subsection{Evidence tiers}
\label{sec:stat-rigor-addendum:tiers}

Let $s_g$ denote the number of distinct seeds for a
(framework, model, task, algorithm) group $g$, and let
$\tau_g = \min_{i \in g} \mathrm{steps}_i$ denote the minimum training
horizon within the group. We define
\begin{equation}
  \label{eq:tiers}
  \mathrm{tier}(g)
  =
  \begin{cases}
    A & s_g \ge 5 \;\text{and}\; \tau_g \ge 100, \\
    B & 3 \le s_g \le 4 \;\text{and}\; 50 \le \tau_g < 100, \\
    C & \text{otherwise.}
  \end{cases}
\end{equation}
Tier-A groups carry enough replication and horizon to support a
two-sample Welch test with the minimum detectable effect size
reported in Section~5 (MDE $d_{\min} \approx 2.02$ for $n_1=n_2=5$,
$\alpha{=}0.05$, $1-\beta{=}0.80$); they are our \emph{inferential}
tier. Tier-B groups support \emph{supporting} evidence: CIs and
point estimates that we are willing to state, but not as BH-corrected
significance claims on their own. Tier-C groups -- including
\emph{all} Tinker API runs, the partial Modal Qwen3-32B PPO run, and
every frontier MoE checkpoint -- are strictly \emph{descriptive}:
we retain them as case studies but they are excluded from the BH
$p$-value family.

\subsection{BH correction after Tier-C exclusion (W5)}
\label{sec:stat-rigor-addendum:bh}

The paper-wide $p$-value family in
Tables~\ref{tab:main_results_stats}--\ref{tab:ppo_grpo_stats} is
\emph{$k=38$} tests, of which $k_A + k_B$ are Tier-A or Tier-B and the
remainder are Tier-C entries with $n_1{=}n_2{=}1$ (i.e.\ no within-group
variance and no valid Welch $t$-statistic). Including those rows in
the BH denominator forces $p^{(i)}_{\mathrm{BH}}
= p^{(i)} \cdot m / \mathrm{rank}(p^{(i)})$ with an artificially
inflated $m$, which for Tier-A findings is strictly anti-conservative
after we drop the uninformative Tier-C rows.

Concretely, let $\mathcal{F}_{AB} \subseteq \{1,\ldots,38\}$ be the
restricted family of Tier-A/B tests, and let $m' = |\mathcal{F}_{AB}|$.
We re-run BH~\citep{benjamini1995controlling} with FDR $q{=}0.05$ on
$\mathcal{F}_{AB}$ only. A test $i \in \mathcal{F}_{AB}$ survives the
corrected family iff
\begin{equation}
  \label{eq:bh-restricted}
  p^{(i)} \le \frac{\mathrm{rank}(p^{(i)})}{m'}\cdot q,
\end{equation}
where ranks are taken within $\mathcal{F}_{AB}$. This is the procedure
used throughout this appendix. Tier-C rows retain their raw
$p$-values where defined, but are labelled ``not corrected'' in the
exported TSV.

\subsection{Survival table for $F_1$--$F_5$ (W4, Q5)}
\label{sec:stat-rigor-addendum:survival}

Table~\ref{tab:survival-f1-f5} shows, for each headline finding,
whether Tier-A evidence exists, whether Tier-B evidence exists,
the Cohen's $d$ of the Tier-A/B restricted comparison, the
BH-adjusted $p$-value under the restricted family of
\S\ref{sec:stat-rigor-addendum:bh}, and our current verdict. The table
is regenerated directly from
\texttt{experiments/results/survival\_analysis.tsv}.

\begin{table}[ht]
\centering
\caption{\textbf{Survival of $F_1$--$F_5$ after Tier-C exclusion.}
``Tier-A'' requires $\ge 5$ seeds and $\ge 100$ steps in a single open
framework (TRL); ``Tier-B'' requires $\ge 3$ seeds and $\ge 50$ steps.
BH correction applied only over Tier-A/B $p$-values. ``Survives''
means the BH-adjusted $p$ is below $q{=}0.05$ \emph{and} the effect
size is at least small ($|d| \ge 0.3$). Findings that do not survive
are \emph{downgraded to descriptive} in the main text (W4).}
\label{tab:survival-f1-f5}
\small
\begin{tabular}{@{}l l c c r r l@{}}
\toprule
\textbf{Finding} & \textbf{Claim (short)}
  & \textbf{Tier-A?} & \textbf{Tier-B?}
  & $d$ & $p_{\mathrm{BH}}$ & \textbf{Survival} \\
\midrule
$F_1$ & ZVF diagnostic              & no  & no  & --     & --       & downgraded (Tier-C) \\
$F_2$ & Instruct $>$ Base            & no  & no  & --     & --       & downgraded (Tier-C) \\
$F_3$ & PPO/GRPO heterogeneity       & yes & no  & $+22.44$ & $6.7\times 10^{-11}$ & \textbf{survives} \\
$F_4$ & Framework gap                & no  & no  & --     & --       & downgraded (Tier-C) \\
$F_5$ & Frontier stability           & no  & no  & --     & --       & downgraded (Tier-C) \\
\bottomrule
\end{tabular}
\end{table}

The table is deliberately conservative. In the current release only
$F_3$ (the large PPO-vs.-GRPO / classic-RL gap, driven by
$n_1{=}n_2{=}5$ arithmetic seeds on TRL, SB3, CleanRL and Tianshou)
has enough Tier-A replication in a single open framework to pass the
restricted BH test. $F_1$ (ZVF as a diagnostic) is supported only by
short-horizon Tinker logs and hence cannot be tested at Tier A;
$F_2$ (instruct $>$ base trainability) is observed across Tinker
runs but we do not yet have paired $\ge 5$-seed $\ge 100$-step
instruct/base comparisons in an open framework; $F_4$ (the framework
gap on Qwen3-8B) is a single-seed four-way comparison on Tinker,
TRL, veRL and OpenRLHF with $s_g{=}1$ per cell and is therefore
Tier-C; $F_5$ (frontier-model stability) is every single-seed MoE
case study combined, which by construction cannot be Tier-A.

\subsection{Claims that are explicitly downgraded (W4)}
\label{sec:stat-rigor-addendum:downgrades}

To make the consequences of Tier-C exclusion concrete, we list the
specific main-text claims whose evidentiary status we revise:

\begin{enumerate}
  \item \textbf{Frontier model training reward.} The Tinker API
    rows for Qwen3-32B, Qwen3.5-27B, Nemotron-120B, Qwen3-235B-A22B,
    Qwen3-30B-A3B (both variants), Kimi-K2 / Kimi-K2-Thinking,
    GPT-OSS-120B and DeepSeek-V3.1 in Table~\ref{tab:main_results}
    are Tier-C. They remain in the table as case studies; we no
    longer attach BH-adjusted $p$-values to them and we strike the
    word ``significant'' from their textual discussion.
  \item \textbf{Framework gap (Task-4 row block in
    Table~\ref{tab:main_results}).} The Tinker / TRL / veRL /
    OpenRLHF four-bar comparison on Qwen3-8B is $s_g{=}1$ per cell
    and therefore Tier-C. The 17$\times$ ratio between Tinker and
    TRL last-10 reward is reported descriptively but is not a
    tested effect.
  \item \textbf{PPO vs.\ GRPO on Qwen3-8B.} The single-seed rows
    $0.344$ (GRPO) and $0.350$ (PPO) do not enter the BH family.
    The paired Welch $p$ of $0.973$ in
    Table~\ref{tab:ppo_grpo_stats} is preserved in the table for
    reference but demoted to ``descriptive'' in the caption.
  \item \textbf{Frontier stability proxy (SI/PTD).} The per-run SI
    and PTD values for all Tinker MoE runs are Tier-C and are no
    longer counted as observations in the paper-wide BH family.
\end{enumerate}

The negative held-out GSM8K result (Qwen3-8B post-GRPO $83.3\%$ vs.\
base $82.0\%$, $p{=}0.26$) is unaffected: that evaluation is
$5$-seed on TRL with a common eval harness and is Tier-A; if
anything, its non-significance is \emph{strengthened} by restricting
to the Tier-A/B family.

\subsection{Reproducibility artifact}
\label{sec:stat-rigor-addendum:repro}

All numbers in Table~\ref{tab:survival-f1-f5} are deterministic.
The driver is
\texttt{experiments/survival\_analysis.py}; it reads
\texttt{experiments/results/*.jsonl},
\texttt{experiments/results/**/*.jsonl},
\texttt{experiments/master\_results.json} and
\texttt{experiments/master\_results.csv}, assigns tiers per
Eq.~\eqref{eq:tiers}, applies BH per Eq.~\eqref{eq:bh-restricted}
over Tier-A/B only, and writes
\texttt{experiments/results/survival\_analysis.tsv} with the
columns \texttt{finding}, \texttt{tier\_a\_support},
\texttt{tier\_b\_support}, \texttt{effect\_size\_cohens\_d},
\texttt{bootstrap\_ci\_low}, \texttt{bootstrap\_ci\_high},
\texttt{n\_runs\_used}, \texttt{bh\_adjusted\_p}, and
\texttt{conclusion}. If no Tier-A groups are present, the script
degrades gracefully, emits a schema-only TSV and prints a stderr
warning rather than raising; this is the expected behaviour on
restricted anonymization snapshots that ship without the $5$-seed
TRL runs. The seed for the bootstrap
(\texttt{MASTER\_SEED}~$=~20260506$) matches
\texttt{experiments/compute\_statistics.py}, so both pipelines are
bit-reproducible on the committed artefacts.

\paragraph{Summary.}
After excluding single-seed short-horizon runs from the inferential
family, only $F_3$ survives the restricted BH correction in the
current release. $F_1$, $F_2$, $F_4$ and $F_5$ are downgraded to
descriptive case studies until matched multi-seed $\ge 100$-step
Tier-A evidence is collected in an open framework -- exactly the
next-step programme called out in the Conclusion.
