\chapter*{Comprehensive Experiment and Results Coverage}
\addcontentsline{toc}{chapter}{Comprehensive Experiment and Results Coverage}

The embedded base report, \texttt{group6.pdf}, is the Group 6 capstone submission
circulated for plagiarism checking on April 5, 2026. Since that submission, the
repository accumulated a substantially broader experiment record spanning later
ablations, held-out evaluation, frontier-model runs, and negative results. This
addendum consolidates that later evidence so the merged final report reflects the
complete project state as of April 20, 2026.

Across the repository we now record 77+ Tinker runs, including a 32-run
``10x Structural Ceiling'' sweep, 13 completed or partially completed GRPO runs
in the World-Class Suite, 2 completed PPO baselines on Modal H100, 1 failed
KL-tracking run, a 5-seed TRL baseline, 5-seed held-out GSM8K evaluation, and
multiple LoRA-rank, group-size, learning-rate, and synthetic-versus-real-data
ablations. The authoritative detailed ledger remains
\path{reports/final/capstone_final_report.md}, especially Sections 4, 5, and 8;
the purpose of this chapter is to surface that evidence inside the final merged
PDF rather than leaving it implicit in repository files.

\section*{Tool Calling, Structured Output, and Code Generation}

The tool-use track now covers four distinct lines of evidence: Sandhya's
real-dataset SFT$\rightarrow$GRPO scaling study, Arumugam's independent
replication, Dhruva's schema-compliance pipeline, and Arvind's Tinker-based
tool-use GRPO runs on both synthetic and real tool corpora. Together these
experiments establish three points clearly. First, tiny synthetic datasets are
insufficient: Sandhya's 0.5B SFT-only run on 14 examples remained largely
non-functional. Second, once the model is warm-started with SFT on a real tool
distribution, GRPO sharply improves structured calling behavior. Third, the
apparent ease of synthetic tool benchmarks does not transfer to real tool
schema diversity.

\small
\begin{center}
\begin{tabularx}{\textwidth}{L{2.3cm}L{3.3cm}L{3.0cm}Y}
\toprule
Study & Setup & Main quantitative result & What it establishes \\
\midrule
Sandhya Exp.\ 1 & Qwen2.5-0.5B-Instruct, 14 synthetic examples, SFT only & JSON valid 30\%, correct tool 20\%, full match 0\% & A tiny handcrafted set is not enough to induce usable tool calling. \\
Sandhya Exp.\ 2 & Qwen2.5-1.5B-Instruct, Glaive-function-calling-v2 plus ToolBench, SFT then GRPO & JSON valid 0\%\,$\rightarrow$\,92\%, correct tool 0\%\,$\rightarrow$\,50\%, arguments 0\%\,$\rightarrow$\,42\%, clean output 0\%\,$\rightarrow$\,92\%, average score 0.0\,$\rightarrow$\,0.59 & Real-data warm-start plus GRPO creates structured tool calls where SFT alone produced plain text. \\
Sandhya Exp.\ 3 & Qwen2.5-3B-Instruct, multi-turn tool chaining, 40 GRPO steps & Mean score 0.72\,$\rightarrow$\,0.91 & Multi-turn chaining improves strongly under GRPO once repeat-call penalties are added. \\
Arumugam & Independent replication of Sandhya Exp.\ 2 & JSON 0\%\,$\rightarrow$\,92\%, correct tool 0\%\,$\rightarrow$\,50\%, average 0\,$\rightarrow$\,0.59 & The core real-data single-turn tool result reproduces independently. \\
Dhruva & Synthetic five-tool benchmark plus SFT+GRPO schema pipeline & Schema compliance 52\%\,$\rightarrow$\,70--71\% in early phases and 97\%\,$\rightarrow$\,100\% at the final stage & GRPO is effective for near-ceiling format refinement, even when held-out task quality does not move. \\
Arvind tool-use Tinker & Qwen3-8B, 30-step and 100-step runs on synthetic 5-tool data, xlam-60k, and math & 30-step synthetic last-10 rewards 0.875, 0.999, 0.977; 30-step xlam-60k 0.363; 100-step synthetic 0.825; 100-step xlam-60k 0.113; 100-step math 0.264 & Real-world tool corpora are 3--8$\times$ harder than clean synthetic tool tasks. \\
Madhu code generation & Qwen3-8B, 3,000 SFT examples plus 35 GRPO prompts with 10 rollouts each & HumanEval pass@1 reaches 86\% (141/164); earlier 50-problem subset and prior SWE run are superseded & The strongest code-generation result is on full HumanEval, not on the earlier smaller subset. \\
\bottomrule
\end{tabularx}
\end{center}
\normalsize

These results also clarify two negative findings that matter for the final
story. The first is that GRPO without SFT warm-up does not bootstrap tool use:
later World-Class cross-tool runs on Qwen3-32B and Llama-3.1-8B-Instruct both
stayed at exactly 0.0 reward for all 30 steps. The second is that synthetic
tool data materially overstates real performance: the repository's xlam-60k
results remain far below the synthetic five-tool sandbox despite using the same
training machinery.

\section*{Mathematical Reasoning, Held-Out Evaluation, and Ablations}

The GSM8K track is now broad enough to separate three different claims that the
earlier report could not cleanly separate: behavior during stochastic training,
generalization under greedy held-out decoding, and how much of the held-out
score comes from the base model rather than from GRPO. The project also now
contains multi-seed replication, Qwen3.5-4B replication, LoRA-rank ablations,
group-size ablations, a lower-learning-rate control, and a TRL baseline.

\small
\begin{center}
\begin{tabularx}{\textwidth}{L{2.7cm}L{3.3cm}L{2.6cm}Y}
\toprule
Experiment family & Scope & Main quantitative result & Interpretation \\
\midrule
Qwen3-8B multi-seed GRPO & 5 seeds, 50 steps, rank 32 & Mean last-10 reward 30.5\% $\pm$ 3.3\%, 95\% CI [26.5\%, 34.5\%], mean peak 72.5\% & Training trajectories are noisy, but the final last-10 reward is reproducible across seeds. \\
Qwen3.5-4B replication & 4 seeds, 50 steps, rank 32 & Mean last-10 reward 84.7\%, SD 12.0\%, all seeds peak at 100\% & The 4B instruct model is dramatically stronger than the Qwen3-8B setup used in the original campaign. \\
LoRA rank ablation & Ranks 8, 16, 32, 64 on Qwen3-8B & Rank 64 starts fastest (47.5\% first-5) and peaks highest (87.5\%); rank 32 remains the default multi-seed setting & More adapter capacity accelerates early learning, but does not by itself remove the task ceiling. \\
Group-size ablation & $G=4,8,16,32$ on Qwen3-8B & Last-10 reward rises from 23.8\% at $G=4$ to 54.7\% at $G=32$; $G=32$ reaches 100\% peak and 0\% zero-reward groups & Exploration matters strongly; larger groups reduce wasted all-equal-reward batches. \\
Low-LR extended run & Qwen3-8B, 100 steps, LR $5\times10^{-6}$ & Peak 75.0\%, last-10 27.5\% & Lower LR stabilizes but does not break the underlying reward-sparsity ceiling. \\
Held-out GSM8K evaluation & 5 seeds, 200 test items per seed, greedy decoding & Mean held-out accuracy 83.3\%, SD 2.2\%, CI [80.6\%, 86.0\%] & Final checkpoints evaluate well on held-out greedy decoding. \\
Base-model control & Same 200 held-out items, Qwen3-8B without LoRA & 82.0\% held-out accuracy, delta +1.3pp relative to GRPO mean, $t=1.32$, $p=0.26$ & The held-out score is overwhelmingly inherited from the base model; the GRPO gain is not statistically significant. \\
TRL baseline & Qwen2.5-0.5B, 5 seeds, 125 steps on NVIDIA L4 & Mean 73.4\% $\pm$ 7.03\%, bootstrap CI [67.9\%, 78.9\%], versus Tinker comparison $t=8.44$, $p=0.0014$ & Framework and scale matter strongly; the low-cost reference implementation underperforms the Tinker-based frontier setups. \\
\bottomrule
\end{tabularx}
\end{center}
\normalsize

This is the most important correction to the earlier narrative: the repository
now demonstrates that training reward and held-out accuracy answer different
questions. The 30.5\% mean last-10 training reward for the five-seed Qwen3-8B
campaign does \emph{not} imply poor final decoding quality; conversely, the
83.3\% held-out score does \emph{not} imply that GRPO created that capability.
The base-model control shows that most of the held-out GSM8K accuracy was
already present before RL, and the repository now records that as a negative
result rather than overstating the GRPO contribution.

\section*{10x Structural Ceiling Sweep}

The dedicated 10x Structural Ceiling sweep is one of the strongest additions
after the original final report. This 32-run Tinker campaign, costing roughly
\$65, systematically ablated benchmark type, architecture family, model scale,
group size, learning rate, and constrained decoding under a consistent 50-step
training budget.

\small
\begin{center}
\begin{tabularx}{\textwidth}{L{2.5cm}L{3.0cm}L{2.6cm}Y}
\toprule
Axis & Conditions covered & Main result & Conclusion \\
\midrule
Benchmark hierarchy & Tool-use, GSM8K, MATH-500, HumanEval on Qwen3-8B & Tool-use 1.000, GSM8K 0.972--1.000, MATH-500 0.574, HumanEval 0.000 & GRPO solves structural-format tasks, partly handles math, and completely fails on the code benchmark used here. \\
Cross-family dependence & Qwen3-8B, Llama-3.1-8B-Instruct, Llama-3.1-8B base & Tool-use: 1.000 vs 0.103 vs 0.000; GSM8K: 1.000 vs 0.969 vs 0.047 & Tool-use trainability is strongly family-dependent; instruction tuning is a prerequisite. \\
Model-size ladder & Qwen 0.6B, 1.7B, 8B and Llama 1B, 3B controls & Sub-8B instruct-free models remain at or near zero with onset at step 0 and ZVF above 88\% & Below the 8B-instruct regime, groups often never produce reward variance at all. \\
Group saturation metric & $G=4,16,32,64$ & $G=32$ gives mean GU 54.5\% with saturation onset at step 29 & Zero-Variance Fraction and Gradient Utilization are useful training-health diagnostics. \\
Learning-rate tradeoff & $10^{-5}$, $4\times10^{-5}$, $10^{-4}$, $3\times10^{-4}$ & LR $10^{-5}$ never saturates; LR $3\times10^{-4}$ recovers by 50 steps; LR $10^{-4}$ saturates fastest & Speed, saturation, and stability trade off against each other; early partial traces can be misleading. \\
Constrained decoding & Constrained vs unconstrained tool-use training & 0.981 vs 0.998 final reward, same onset step & Grammar-constrained decoding is not the reason the tool-use result succeeds. \\
Reward-hacking collapse & Llama-3.1-8B base on tool-use & Breakout to 0.873 followed by collapse to 0.000 at step 41 & Catastrophic collapse is directly observed in at least one base-model tool-use run. \\
\bottomrule
\end{tabularx}
\end{center}
\normalsize

This sweep turns several earlier intuitions into explicit experimental claims:
benchmark difficulty is sharply ordered, instruction tuning matters more than
raw parameter count for trainability, and the absence of within-group reward
variance is a concrete failure mode rather than a vague stability problem.

\section*{World-Class Suite, PPO Baselines, and Frontier Runs}

The World-Class Suite extends the project beyond the original 0.5B--8B regime
to large dense and MoE checkpoints, and it also adds the first PPO baselines on
Modal H100. The current ledger includes 13 completed or partially completed GRPO
runs on Tinker, 2 completed PPO baselines, 1 failed KL-instrumentation run, and
1 GPT-OSS-20B run that did not complete because of inference-side Tinker API
timeouts.

\small
\begin{center}
\begin{tabularx}{\textwidth}{L{2.8cm}L{1.5cm}L{1.1cm}L{1.3cm}Y}
\toprule
Model & Arch / scale & Steps & Peak / last-10 & Coverage note \\
\midrule
DeepSeek-V3.1 & MoE, about 671B total, 37B active & 20 & 100\% / 85.0\% & High reward from step 1; capability largely pre-exists RL. \\
Qwen3.5-4B & Dense 4B & 30 & 100\% / 85.0\% & Strong completed GRPO run in the smaller-model regime. \\
Llama-3.1-8B-Instruct & Dense 8B & 30 & 100\% / 84.4\% & Strong GRPO baseline later compared against PPO. \\
Qwen3-8B & Dense 8B & 30 & 62.5\% / 34.4\% & Much weaker than the best 50-step Qwen runs; shows trajectory sensitivity. \\
Qwen3-235B-A22B & MoE, 235B total, 22B active & 15 & 100\% / 100\% & Fastest perfect convergence in the whole study. \\
Qwen3-30B-A3B-Instruct & MoE, 30B total, about 3B active & partial & 100\% / 100\% & Partial run, but already at perfect reward. \\
Nemotron-120B & Dense 120B & 20 & 87.5\% / 16.2\% & Peak-then-collapse behavior at frontier scale. \\
Qwen3.5-27B & Dense 27B & 10 & 75.0\% / 43.7\% & Partial ascending run, suggests more room with longer training. \\
Qwen3-32B & Dense 32B & 10 & 31.2\% / 25.0\% & Slow early ascent under the available partial budget. \\
Qwen3-30B-A3B (base) & MoE base & partial & 50.0\% / 32.5\% & Same architecture as instruct variant, showing the effect of instruction tuning. \\
Kimi-K2 & MoE, about 1T total & 20 & 100\% / 80.0\% & Strong frontier MoE result with moderate late instability. \\
\bottomrule
\end{tabularx}
\end{center}
\normalsize

\small
\begin{center}
\begin{tabularx}{\textwidth}{L{3.2cm}L{2.6cm}L{2.0cm}Y}
\toprule
Additional run & Scope & Main result & What it means \\
\midrule
Cross-tool Qwen3-32B & Tool-use GRPO without SFT warm-up & 0\% peak, 0\% last-10 & Tool-use never bootstraps without a warm-started instruct policy. \\
Cross-tool Llama-3.1-8B-Instruct & Tool-use GRPO without SFT warm-up & 0\% peak, 0\% last-10 & The no-SFT failure is not unique to Qwen. \\
PPO Llama-3.1-8B-Instruct & Modal H100 PPO baseline & 100\% peak, 97.5\% last-10 & Strongest late-stage stability in the entire record. \\
PPO Qwen3-8B & Modal H100 PPO baseline & 100\% peak, 22.5\% last-10 & Same method behaves very differently on Qwen than on Llama. \\
KL tracking run & Modal H100 instrumentation & final\_kl 60.75, trajectory unusable & Gradient-graph bug prevented reliable KL and entropy logging. \\
GPT-OSS-20B & Planned frontier GRPO run & did not complete & Tinker inference timeouts blocked the run before usable results were obtained. \\
\bottomrule
\end{tabularx}
\end{center}
\normalsize

The main conclusion from this suite is not a simplistic ``bigger is better''
story. The frontier results are non-monotonic: some very large models converge
immediately, some show strong but unstable performance, and one 120B dense
model collapses badly. The clearer pattern is that instruction tuning and model
family dominate trainability. That is why Qwen3-30B-A3B-Instruct reaches perfect
reward while the otherwise matched base MoE checkpoint remains far lower, and it
is why PPO helps Llama dramatically but destabilizes Qwen.

\section*{Negative Results, Null Findings, and Scope Corrections}

The final report should preserve the failures as explicitly as the successes.
The repository now supports the following negative or limiting conclusions.

\begin{itemize}
\item The held-out GSM8K improvement over the base Qwen3-8B model is only
+1.3 percentage points and is not statistically significant ($p=0.26$).
\item Tool-use GRPO without SFT warm-up failed completely on both
Qwen3-32B and Llama-3.1-8B-Instruct, staying at exactly 0.0 reward.
\item The 10x Structural Ceiling sweep produced a total null on the HumanEval
code benchmark under the tested GRPO setup.
\item Nemotron-120B is a frontier-scale counterexample to any simple
``frontier models are always stable'' narrative: it peaks high and then
collapses to a poor last-10 average.
\item KL and entropy tracking are not yet a finished result. The PPO reward
curves are valid, but the KL instrumentation run failed because the loss term
was not properly attached to the autograd graph.
\item GPT-OSS-20B is an incomplete run, not a weak result. The blocker was
Tinker-side inference timeout, not a measured training failure.
\item Earlier smaller code-evaluation claims were superseded by the final full
HumanEval evaluation and should no longer be treated as the definitive result.
\end{itemize}

\section*{Reproducibility Ledger}

The merged PDF now contains the experiment-level conclusions, but the detailed
artifacts still live in the repository and should be treated as the canonical
evidence trail:

\begin{itemize}
\item \path{reports/final/capstone_final_report.md}: authoritative narrative,
run registry, statistical summary, platform and budget breakdown, and links to
checkpoints and W\&B.
\item \path{reports/final/supplementary_appendix.tex}: paper appendix material,
including reviewer-facing statistical clarifications and surviving claims.
\item \path{paper/main.tex} and the appended \path{paper/main.pdf}: final paper
text integrating the later analysis, reviewer responses, and formalized
results.
\item \path{experiments/results/} and the JSON files under
\path{reports/final/}: held-out GSM8K outputs, baseline controls, and generated
diagnostic tables.
\item HuggingFace checkpoints under \url{https://huggingface.co/arvindcr4} and
Weights \& Biases projects \texttt{tinker-structural-ceiling} and
\texttt{tinker-rl-lab-world-class}: external storage for run traces and model
artifacts referenced throughout the capstone report.
\end{itemize}

This chapter is therefore the bridge between the original submitted report and
the much larger body of evidence accumulated afterward: the embedded base report
preserves the submitted document, while this addendum records the full later
experiment corpus and the appended paper preserves the final research framing.
